Labor Market Income and Substitution Effects Explain Modern Natural Selection

### 2020-12-10 why-negative-selection.Rmd
Use residualized scores as the default. Add materials and methods section. Final tweaks for first draft.

# nasty hack to make add_ashe_income work later
source("~/import-ukbb-data/import-ukbb-data.R")

@@

@@

# Data and methods

We use data from UK Biobank. Our sample consists of white British (as defined
by genetic estimated ancestry). For regressions on number of children we 
limit the sample to people over 45, since most fertility is completed by this age. We compute 33 polygenic scores, normalized
to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order
of 5% of a standard deviation.

@@

These changes could reflect natural selection within the UK
population, but also ascertainment bias within the sample. Respondents are
higher income and better educated than the UK population, and they may
also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias might also
increase with age.

@@

The "selection effect", $\beta$, measures the strength of natural
selection within the sample. In fact, since polygenic scores are
normalized, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic].^[normalization] To learn more about the
underlying mechanisms, we split the sample, starting with basic
demographic variables including education, income and sex. These are all
potential sources of ascertainment bias: as well as the ascertainment
for income and education, mentioned above, the sample sex ratio skews
`r female_ratio * 100`% female.

^[normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
where $Y$ is the number of children. Since PGS are normalized to variance 1,
this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted
by the number of children, which is the average polygenic score in the 
next generation.

@@

@@

res_wt_combined <- bind_rows(
  None                     = res_all %>% 
                               filter(reg.type == "raw", dep.var == "N children") %>% 
                               select(-reg.type, -dep.var, -score_name),
  Geographical             = res_wt_msoa_weights,
  `Age/Qualification`      = res_wt_age_qual_weights,
  `Age/Qual/AFLB`  = res_wt_flb_weights,
  .id = "Weights"
)

@@

Since UK Biobank subjects are not representative of the wider population, our
results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using data from
the 2011 Census and from the 2006 General Household Survey (GHS). We try three
alternative weighting schemes: geographical weighting by census MSOA, age and
presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Mean effect sizes across all polygenic scores are increased by a
factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the Biobank
sample seems to be ascertained in ways that shrink estimates of selection
effects, we suspect that our weighted estimates are still conservative.

@@

Although the direction of selection effects does not change between the generations,
there are other differences. After we residualize polygenic scores on the first
100 principal components of the array data, selection effects on *number of
siblings* are reduced by about ten percent on average, whereas effects on
*number of children* barely change (appendix Figure
\@ref(fig:plot-regs-controlled-pcs)). This could be because
earlier fertility is driven more by geographically clustered deprivation [e.g.
via an insurance motive, @rendall1998old], which may correlate with the
broad-scale genetic variation captured by principal components. Another
difference is that in the parents' generation, raw selection effects and effects
controlling for AFLB are positively correlated, whereas they are negatively
correlated in the respondents' generation.[^educ-exception] These differences
suggest that the mechanisms of natural selection may have changed over time,
even though the outcomes are broadly the same.

[^educ-exception]: Educational attainment is an exception. It is selected 
against in the parents' generation, but selected for if we control for AFLB,
which is the same pattern we observe among the respondents' generation.

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by area Townsend deprivation quintile", fig.align = "center"}

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (Figure \@ref(fig:plot-siblings-townsend))
show the same pattern as for respondents: effect sizes are larger and more
often significant in the most deprived areas.

@@

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is `r sib_chn_cor`.

# Economic theory and natural selection

@@

The theory can also explain why natural selection is weaker among higher-income
parents. @becker1976child show that if parents care about child quality as well
as quantity, under certain assumptions the income effect will be stronger at
higher income levels; this can lead to a U-shaped relationship between income
and fertility. Empirically, @cohen2013financial find that income decreases 
fertility at low income levels but increases it at higher income levels.

@@

effect_size <-  res_all %>% 
                  filter(dep.var == "N children", reg.type == "raw") %>% 
                  pull(estimate)

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None"),
        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.012) +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        coord_cartesian(clip = "off") +
        theme(
          legend.position = "none",
          panel.spacing = unit(14, "pt")
        )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each PGS against that PGS's correlation with earnings in a respondent's first
job[^earnings-est], and against its correlation with educational attainment, a
predictor of lifetime earnings. The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
Earnings, using the SOC 2000 job code.

@@

res_compare <- res_all %>% 
  filter(dep.var=='N children', reg.type == 'raw') %>% 
  left_join(res_risk_control %>% filter(term == score_name), by = 'score_name')

ratio <- res_compare$estimate.y/res_compare$estimate.x
n_less_sig <- sum(res_compare$p.value.y <= 0.05/33 & res_compare$p.value.x > 0.05/33)
```

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the PGS. The median ratio of PGS effect sizes
between regressions with and without controls is `r median(ratio)`; all PGS
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the leading explanation.

@@

Previous work has documented natural selection in modern populations on variants
underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that natural selection effects are highly
concentrated among specific subgroups of the population, including poorer
people, less educated people, younger parents and those with more lifetime
sexual partners. Indeed, among older parents selection effects are actually
reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. The 
economic theory of fertility provides a parsimonious explanation for this.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In short, it is probably too early
to tell whether modern natural selection has a substantively important effect on
population genetics. Nevertheless, we note that selection effects on our measured
polygenic scores are still relatively small, even after reweighting to account for ascertainment bias.

@@

\FloatBarrier
# Appendix

@@

Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire
sample, using our three weighting schemes. We include unweighted regressions
for comparison. Table \@ref(tab:tbl-res-wt) gives effect sizes as a
proportion of the unweighted effect size.

@@

rel_sizes %>% 
      arrange(desc(Geographical)) %>% 
      bind_rows(rel_sizes_summary) %>% 
      rename(PGS = term) %>% 
      as_hux() %>% 
      insert_row("", "Weighting", fill = "") %>% 
      set_caption(
        "Weighted effect sizes as a proportion of unweighted"
      ) %>% 
      merge_cells(1, 2:4) %>% 
      set_align(1, 2, "centre") %>% 
      set_header_rows(1, TRUE) %>% 
      theme_compact() %>% 
      style_headers(bold = TRUE) %>% 
      set_bottom_border(1, 2:4) %>% 
      set_bottom_border(2, everywhere) %>% 
      set_italic(final(2), 1) %>% 
      set_font_size(8) %>% 
      set_width(0.7) %>% # ensures line wrapping below
      set_col_width(c(.34, .22, .22, .22)) %>% 
      add_footnote(paste0(
        "Only consistent and significant (when unweighted) estimates ",
        "are shown.\n",
        "Age/Qual/AFLB as a proportion of unweighted regressions ",
        "for females only."),
        font_size    = 6, 
        top_padding  = 0,
        left_padding = 0
      )

```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions"}

@@

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. For comparison, we show regressions on the respondents'
generation (*number of children*), where respondents with no
children have been excluded.

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings)")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on number of children)")
```

\FloatBarrier
## Controlling for principal components

Polygenic scores could capture effects that are really due to population
stratification, although would not change our results for natural
selection of the scores. Figure \@ref(fig:plot-regs-controlled-pcs) shows
selection effects on polygenic scores residualized for the top 100
principal components of the genetic data, calculated within the UK
Biobank population.

@@

In siblings regressions, effect sizes are smaller when residualizing for
principal components -- sometimes much smaller, as in the case of
height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled"
effect sizes have a smaller absolute value than the corresponding "raw"
effect size. The median proportion between raw and controlled effect
sizes is `r median_prop_sibs`. Among the children regressions, this no
longer holds. Effect sizes are barely affected by controlling for
principal components.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using residualized polygenic scores on number of siblings/children."}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free_y") +
        pval_geom_point(position = position_dodge(0.4)) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered()

@@

To get a further insight into this we regress *siblings* and *children*
on individual principal components. As Figure \@ref(fig:plot-pc-regs)
shows, effects are larger and more significant in siblings regressions.
`r prop_sig[["N siblings"]]` principal components significantly
predicted number of siblings, while only `r prop_sig[["N children"]]`
significantly predicted number of children.

@@

res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
compare_abp_raw <- res_age_birth_parents %>% 
      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
      left_join(res_sibs_raw, by = "term")

cor_raw_fath <- with(
  compare_abp_raw %>% filter(control == "fath_age_birth"), 
  cor(estimate.x, estimate.y))
cor_raw_moth <- with(
  compare_abp_raw %>% filter(control == "moth_age_birth"), 
  cor(estimate.x, estimate.y))
```

Among the parents' generation, we can control for age at first live
birth using the subsets of respondents who reported their mother's or
father's age, and who had no elder siblings. We run regressions on
*number of siblings* on these subsets, controlling for either parent's
age at their birth. Figure \@ref(fig:plot-age-birth-parents) shows the
results. Effect sizes are very similar, whether controlling for father's
or mother's age. As in the respondents' generation, effect sizes
are negatively correlated with the effect sizes from bivariate
regressions without the age at birth control, though the correlation
is smaller (father's age at birth: $\rho$ `r cor_raw_fath`; mother's age at
birth: $\rho$ `r cor_raw_moth`).

@@

res_all %<>% 
  filter(term == score_name, reg.type == "raw", dep.var == "N children") %>% 
  select(-dep.var, -reg.type)

@@

TODO: ref source of data.

@@

@@

rgs <- left_join(rgs, 
        res_all %>% filter(reg.type == "raw", dep.var == "N children"), 
        by = c("p2" = "term"))

@@

```{r correlate-effect-sizes, eval = FALSE}
combined_results <- res %>% 
      mutate(PCs = ifelse(reg.type=="raw", "no", "yes")) %>% 
      dplyr::select(-reg.type) %>% 
      {bind_rows(bivariate = ., multivariate = all_res, .id = "reg.type")} %>% 
      arrange(term, reg.type, PCs) %>% 
      dplyr::select(term, reg.type, PCs, estimate) %>% 
      pivot_wider(names_from = c(reg.type, PCs), values_from = estimate)

cor(combined_results[-1], use = "complete")

@@

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean polygenic
scores between the highest and lowest income groups. Column "Actual" shows parents' scores weighted by Age/Qualification times number of children. 
Column "Without selection" shows scores weighted by Age/Qualification only, i.e. if all couples had the same number of children.

@@

ineq_comp %>% 
      select(
        PGS                 = score, 
        `With selection`    = diff_children, 
        `Without selection` = diff_hypoth
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(Actual)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group") %>% 
      theme_compact()
```

# References

>>>>

# nasty hack to make add_ashe_income work later:
if (! exists("add_ashe_income")) source("~/import-ukbb-data/import-ukbb-data.R")

@@

comma_num <- function (x) prettyNum(x, big.mark = ",")

@@

# Results

Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in the
sample by 5-year birth intervals. Several scores show consistent increases or
declines over this 30-year period, of the order of 5% of a standard
deviation.These changes could reflect natural selection within the UK
population, but also ascertainment bias within the sample. Respondents are
higher income and better educated than the UK population, and they may
also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias might also
increase with age.

@@

@@

The "selection effect", $\beta$, measures the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] To learn more about the underlying
mechanisms, we split the sample, starting with basic demographic variables
including education, income and sex. These are all potential sources of
ascertainment bias: as well as the ascertainment for income and education,
mentioned above, the sample sex ratio skews `r female_ratio * 100`% female.

[^normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
where $Y$ is the number of children. Since PGS are normalized to variance 1,
this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn
reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted by
the number of children, which is the average polygenic score in the next
generation.

@@

## Correcting for sampling bias

@@

res_wt_combined <- bind_rows(
  None                = res_all %>% 
                          filter(
                            reg.type == "controlled", 
                            dep.var == "N children"
                          ) %>% 
                          select(-reg.type, -dep.var, -score_name),
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

@@

Since UK Biobank subjects are not representative of the wider population, our
results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population data. We try three
alternative weighting schemes: weighting by geography, age and
presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects 
among the entire sample, estimated with the three weighting schemes. 
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the Biobank
sample seems to be ascertained in ways that shrink estimates of selection
effects, we suspect that our weighted estimates are still conservative.

```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions. Unweighted estimates are shown for comparison."}

## Selection in the parents' generation

@@

Although the direction of selection effects does not change between the
generations, there are other differences. Compared to residualized polygenic
scores, selection effects on unresidualized scores are about ten percent higher
on average for *n siblings*, whereas effects for *number of children* barely
change (appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
because earlier fertility is driven more by geographically clustered deprivation
[e.g. via an insurance motive, @rendall1998old], which may correlate with the
broad-scale genetic variation captured by principal components.

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (appendix Figure
\@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
effect sizes are larger and more often significant in the most deprived areas.

@@

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho = `r sib_chn_cor`$).

## Economic fertility theory and natural selection

@@

The theory can also explain why natural selection is weaker among higher-income
parents. @becker1976child show that if parents care about child quality as well
as quantity, under certain assumptions the income effect will be stronger at
higher income levels; this can lead to a U-shaped relationship between income
and fertility. Empirically, @cohen2013financial find that income decreases
fertility at low income levels but increases it at higher income levels. If, in
our data, at high income levels income and substitution effects are roughly
balanced , then income-linked polygenic scores will be neither selected for nor
against.

@@

effect_size <-  res_all %>% 
                  filter(dep.var == "N children", reg.type == "controlled") %>% 
                  pull(estimate)

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None"),
        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.005) +
        facet_grid(Correlation ~ Controls, labeller = label_both,
                   scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "") +
        coord_cartesian(clip = "off") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

## Testing the theory

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each PGS against that PGS's correlation with earnings in a respondent's first
job, and against its correlation with educational attainment, a
predictor of lifetime earnings. The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

res_compare <- res_all %>% 
  filter(dep.var == "N children", reg.type == "controlled") %>% 
  left_join(res_risk_control %>% filter(term == score_name), by = "score_name")

ratio <- res_compare$estimate.y/res_compare$estimate.x
n_less_sig <- sum(res_compare$p.value.y <= 0.05/33 & res_compare$p.value.x > 0.05/33)
stopifnot(n_less_sig == 0)
```

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the PGS. The median ratio of PGS effect sizes
between regressions with and without controls is `r median(ratio)`; all PGS
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the most likely explanation.

@@

Previous work has documented natural selection in modern populations on variants
underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including poorer people, less educated people, younger parents and those with
more lifetime sexual partners. Indeed, among older parents selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In short, it is probably too early
to tell whether modern natural selection has a substantively important effect on
population genetics. Nevertheless, selection effects on our measured
polygenic scores are still relatively small, even after reweighting to 
account for ascertainment bias.

@@

# Materials and methods

We use participant data from UK Biobank [@sudlow2015uk], which has received
ethical approval from the National Health Service North
West Centre for Research Ethics Committee (reference: 11/NW/0382).
We limit the sample to white British participants, as defined by
genetic estimated ancestry and self-identified ethnic group, giving a sample
size of `r comma_num(nrow(famhist))`. For regressions on number of children we use participants over 45, since most fertility is completed by this age. This
gives a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.

We computed 33 polygenic scores. The effect size estimates were obtained from
GWASs that were chosen to not have included the UK Biobank dataset, to avoid
overestimation of the genetic predisposition of a trait. The polygenic scores
were computed using the summary-data based best linear unbiased prediction
(SBLUP) approach. As a reference sample for the linkage disequilibrium, we used
a random sample of 10,000 unrelated individuals from UK Biobank, imputed using
the Haplotype Reference Consortium panel. Scores were normalized to mean 0,
variance 1, and residualized on the first 100 principal components of array
data, which were calculated as in @abdellaoui2019genetic.

Earnings in first job are estimated from mean earnings in the 2007 Annual Survey
of Hours and Earnings, using the SOC 2000 job code (Biobank field 22617).

Population data for weighting is taken from the 2011 UK Census and the 2006
General Household Survey (GHS). Weighting for Age/Qualification and 
Age/Qualification/AFLB weights was done using marginal totals from a linear
model, using the `calibrate()` function in
the R "survey" package [@lumley2020survey]. Geographical weighting was done with
iterative post-stratification using the `rake()` function, on Census Middle Layer
Super Output Areas, sex and presence/absence of a partner.

\FloatBarrier
\newpage
# Appendix

@@

Table \@ref(tab:tbl-res-wt) gives effect sizes as a proportion of the unweighted
effect size, for all polygenic scores which are consistently signed and which
are significantly different from zero in unweighted regressions.

@@

rel_sizes %>% 
      arrange(desc(Geographical)) %>% 
      bind_rows(rel_sizes_summary) %>% 
      rename(PGS = term) %>% 
      as_hux() %>% 
      insert_row("", "Weighting", fill = "") %>% 
      set_caption(
        "Weighted effect sizes as a proportion of unweighted effect sizes."
      ) %>% 
      merge_cells(1, 2:4) %>% 
      set_align(1, 2, "centre") %>% 
      set_header_rows(1, TRUE) %>% 
      theme_compact() %>% 
      set_number_format(-1, -1, 2) %>% 
      style_headers(bold = TRUE) %>% 
      set_bottom_border(1, 2:4) %>% 
      set_bottom_border(2, everywhere) %>% 
      set_italic(final(2), 1) %>% 
      set_font_size(8) %>% 
      set_width(0.7) %>% # ensures line wrapping below
      set_col_width(c(.34, .22, .22, .22)) %>% 
      add_footnote(paste0(
        "Only consistently-signed and significant (when unweighted) ",
        "estimates are shown. ",
        "Age/Qual/AFLB as a proportion of unweighted regressions ",
        "including females only."),
        top_padding  = 0,
        left_padding = 0
      )

@@

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. By definition, people who had no children cannot be included in
this data. For a cleaner comparison, we rerun regressions on the respondents'
generation (*number of children*) after excluding respondents with no
children.

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings).")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on number of children).")
```

Figure \@ref(fig:plot-siblings-townsend) plots effects on *n siblings* by
Townsend deprivation quintile of birth area.

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area", fig.align = "center"}

\FloatBarrier
## Selection effects on raw polygenic scores

@@

Figure \@ref(fig:plot-regs-controlled-pcs) compares
selection effects on polygenic scores residualized for the top 100
principal components of the genetic data, to selection effects on raw, unresidualized polygenic
scores. In siblings regressions, effect sizes are larger for raw scores
principal components -- sometimes much larger, as in the case of
height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "raw"
effect sizes have a larger absolute value than the corresponding "residualized"
effect size. The median proportion between raw and controlled effect
sizes is `r median_prop_sibs`. Among the children regressions, this no
longer holds. Effect sizes are barely affected by controlling for
principal components.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children."}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        reg.type = fct_recode(reg.type, 
                               Residualized   = "controlled", 
                               Unresidualized = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free_y") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x = element_text(size = 8)
        )

@@

To get a further insight into this we regress *n siblings* and *n children*
on individual principal components. As Figure \@ref(fig:plot-pc-regs)
shows, effects are larger and more significant in siblings regressions.
`r prop_sig[["N siblings"]]` principal components significantly
predicted number of siblings, while only `r prop_sig[["N children"]]`
significantly predicted number of children.

@@

res_sibs_unc <- filter(res_all, dep.var == "N siblings", reg.type == "controlled")
compare_abp_unc <- res_age_birth_parents %>% 
      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
      left_join(res_sibs_unc, by = "term")

cor_unc_fath <- with(
  compare_abp_unc %>% filter(control == "fath_age_birth"), 
  cor(estimate.x, estimate.y))
cor_unc_moth <- with(
  compare_abp_unc %>% filter(control == "moth_age_birth"), 
  cor(estimate.x, estimate.y))
```

Among the parents' generation, we can control for age at first live
birth using the subsets of respondents who reported their mother's or
father's age, and who had no elder siblings. We run regressions on
*number of siblings* on these subsets, controlling for either parent's
age at their birth. Figure \@ref(fig:plot-age-birth-parents) shows the
results. Effect sizes are very similar, whether controlling for father's
or mother's age. As in the respondents' generation, effect sizes
are negatively correlated with the effect sizes from bivariate
regressions without the age at birth control, though the correlation
is smaller (father's age at birth: $\rho$ `r cor_unc_fath`; mother's age at
birth: $\rho$ `r cor_unc_moth`).

@@

res_all %<>% 
  filter(term == score_name, reg.type == "controlled", dep.var == "N children") %>% 
  select(-dep.var, -reg.type)

@@

@@

# TODO: ref source of data.

@@

rgs <- left_join(rgs, 
        res_all %>% filter(reg.type == "controlled", dep.var == "N children"), 
        by = c("p2" = "term"))

@@

@@

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
polygenic scores between the highest and lowest income groups. Column "Actual"
shows respondents' scores weighted by Age/Qualification times number of
children. Column "Without selection" shows scores weighted by Age/Qualification
only, i.e. if all couples had the same number of children.

@@

ineq_comp %>% 
      select(
        PGS                 = score, 
        `With selection`    = diff_children, 
        `Without selection` = diff_hypoth
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`With selection`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, -1, 3) %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
      theme_compact()
```

\FloatBarrier
\newpage
# References
